---
author: $\overset{\mathrm{martin.o.berild@ntnu.no}}{10014}$ \and
        $\overset{\mathrm{yaolin.ge@ntnu.no}}{10026}$
date: \today
title: "Project 3 - TMA4315"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
defaultW <- getOption("warn") 

options(warn = -1) 
```

## Problem 1 

We are interested in computing the maximum likelihood and restricted maximum likelihood estimates of the parameters in the linear mixed model 

$$
y_{ij} = \beta_0 + \beta_1x_{ij} + \gamma_{0,i} + \gamma_{1,i}x_{ij} + \epsilon_{ij},
$$
where $\boldsymbol{\gamma}_i = (\gamma_{0,i},\gamma_{1,i})$ are iid binomially distributed with zero mean and covariance matrix

$$
\begin{bmatrix}
\tau_0^2 & \tau_{01} \\
\tau_{01} & \tau_{1}^2
\end{bmatrix},
$$
and $\epsilon_{ij}$ are idd Guassian distributed with zero mean and variance $\sigma^2$ for $i = 1,\dots,m$ and $j=1,\dots,n$. Thus, we are interested in obtaining the estimates of $(\beta_0,\beta_1,\tau_0^2,\tau_1^2,\tau_{01},\sigma^2)$. 


## Problem 2

Next, we are interested in modelling the 2018 results of the Norwegian elite football league using a generalized linear mixed model. First we load the data and display the contents.

```{r dataload}
long <- read.csv("https://www.math.ntnu.no/emner/TMA4315/2020h/eliteserie.csv", colClasses = c("factor","factor","factor","numeric"))
head(long)
```



```{r endup, include=FALSE}
options(warn = defaultW)
```